The hidden manifold distance for functional data

### 2019-09-03 MarieDataAnalysis.tex
wrote section classification of real data

\section{Distance-based functional classification
{[}Marie{]}}\label{distance-based-functional-classification-marie}

In this section, we explore whether our geodesic distance estimator has
benefits for downstream analysis task. There are many tasks we could
consider here such as distance-based nonparametric regression and
distanced-based functional clustering, but we will focus on
distance-based functional classification. It must be noted that while
curve alignment, also known as curve registration, is necessarily
performed as a preprocessing technique prior to clustering and
classification, our geodesic distance estimator allows one to forsake
this step.

For simplicity, assume the task is binary classification. Associated to
each functional object \(x\) is a binary \(y\) indicating class
membership. Consider the classifier proposed in \cite{Ferraty2006} which
is a functional version of the Nadaraya-Watson kernel estimator of class
membership probabilities: \[
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
\] We shall compare our method to using \(L^2\) distance, possibly
weighted, and with curve registration already accomplished. Describe
alternative methods in detail.

The bandwidth in the classifier should be tuned individually for each
method. Also we might need to tune MDS dimension \(s\) since in real
data, the dimension of the manifold might be much higher than
encountered in the simulation scenarios where it never goes above 2.

Datasets used by functional classification papers

\begin{itemize}
\item Wheat, rainfall and phoneme in Aurore's paper "Achieving near-perfect classification for functional data"
\item Berkeley growth curves in \cite{ChenReiss2014}.
\item Tecator and phoneme in \cite{Galeano2015} Mahalanobis technometrics paper.
\item yeast cell cycle gene expression (can't find this publicly) in \cite{LengMuller2005} "Classification using functional data analysis for temporal gene expression data"
\end{itemize}

Datasets used in functional manifold papers

\begin{itemize}
\item Berkeley growth, yeast cell cycle gene expression (can't find this publicly) in \cite{ChenMuller2012}
\item Tecator in \cite{LinYao2017} contamination paper
\item Berkeley growth, gait cycle in \cite{Dimeglio2014} robust isomap paper
\end{itemize}

>>>>

\section{Distance-based functional classification}\label{distance-based-functional-classification-marie}

In this section, we explore whether our geodesic distance estimator has
benefits for downstream analysis task. There are many tasks we could
consider here such as distance-based nonparametric regression and
distanced-based functional clustering, but we will focus on
distance-based functional classification.

We work with the non-parametric classifier proposed in \cite{Ferraty2006} which is a functional version of the Nadaraya-Watson kernel estimator. Given a sample of curves $X_1,\ldots,X_n$ where each function $X_i$ is associated to a class $Y_i$, the estimator of the probability that a new observation $X^\star$ is in class $Y^\star=y$ is defined as
\begin{equation}\label{func_NW}
\hat P(Y^\star = y | X^\star)= \frac{ \sum_{i=1}^n k[h^{-1} d(X^\star,X_i)] {\bf{1}}(Y_i = y) }{ \sum_{i=1}^n k[h^{-1} d(X^\star,X_i)] },
\end{equation}
where $d$ is a distance, $k$ a kernel, $h$ a bandwidth and $\bf{1}$ the indicator function. The curve $X^\star$ is then classified in the class $Y^\star =y$ for which the conditional probability (\ref{func_NW}) is larger. We compare the classification performance obtained with the distance $d$ being equal to our proposed geodesic distance with $s=1,2$ and $3$, and to the one obtained with Spline IsoGeo and Spline L2 described in the simulation study section. We set $k$ to be a quadratic kernel and the bandwidth $h$ is chosen by cross-validation.

The classification task is perform on a famous dataset in Functional Data Analysis which is called the Berkeley growth curves dataset and is available in the \emph{fda} package \cite{Ram-Hoo-Gra} for R \cite{Rproject}. We chose this dataset because we expect the data to contain nonlinear features as it is hinted in \cite{ChenMuller2012}. The data consist in the height of $n=93$ individuals, 39 boys and 54 girls, measured on a common grid $t_1,\ldots,t_K$ of $K=31$ points taken between the ages of 1 and 18 years. The raw data $Y_1,\ldots,Y_n$ are transformed to continuous functions by smoothing with B-spline bases with a roughness penalty on the fourth derivative :
$$\tilde X_i(t) = \sum_{l=1}^{b} c_{il}B_l(t), $$
where
 \begin{eqnarray*}
 (c_{i1},\ldots,c_{ib})&=&\arg \min_{c\in \R^b}\Bigg\{\sum_{j=1}^K\left(\sum_{l=1}^{b} c_{il}B_l(t_j)-Y_{ij}\right)^2\\
 &&\quad +\lambda \left\|\sum_{l=1}^{b} c_{il}\partial^4_tB_l\right\|^2_{L^2}\Bigg\}.
 \end{eqnarray*} 
We use $b=35$ B-splines bases of order 6 and the parameter $\lambda$ is chosen by generalized cross-validation. We then differentiate the resulting fonctions to obtain velocity curves; the data are illustrated for boys and girls on Figure \ref{fig:data_illu}.

To assess the classification performance of the different distances, we randomly split $200$ times the dataset into a training set of $50$ curves and a test set of $43$ curves. For each split, we calculate the misclassification error for the 5 different distances, and we average these errors over the splits. The results are presented in Figure \ref{fig:class_err_velo}, we can see that our method with $s=3$ performs a little bit better than the others and that our method with $s=2$ is equivalent to Spline IsoGeo and Spline L2.

\begin{figure}
\centering
\includegraphics[height=0.55\textheight]{./Velocity_curves_data_illustration.pdf}
\caption{Velocity curves of 39 boys (top) and 54 girls (bottom).}
\label{fig:data_illu}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=0.3\textheight]{./Erreur_classification_velocity.pdf}
\caption{Classification errors for the velocity curves, our method with $s=3$ performs better while our method with $s=2$ is equivalent to Spline IsoGeo and Spline L2.}
\label{fig:class_err_velo}
\end{figure}

% It must be noted that while
% curve alignment, also known as curve registration, is necessarily
% performed as a preprocessing technique prior to clustering and
% classification, our geodesic distance estimator allows one to forsake
% this step.

% For simplicity, assume the task is binary classification. Associated to
% each functional object \(x\) is a binary \(y\) indicating class
% membership. Consider the classifier proposed in \cite{Ferraty2006} which
% is a functional version of the Nadaraya-Watson kernel estimator of class
% membership probabilities: \[
% \hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
% \] We shall compare our method to using \(L^2\) distance, possibly
% weighted, and with curve registration already accomplished. Describe
% alternative methods in detail.
% 
% The bandwidth in the classifier should be tuned individually for each
% method. Also we might need to tune MDS dimension \(s\) since in real
% data, the dimension of the manifold might be much higher than
% encountered in the simulation scenarios where it never goes above 2.

% Datasets used by functional classification papers
% 
% \begin{itemize}
% \item Wheat, rainfall and phoneme in Aurore's paper "Achieving near-perfect classification for functional data"
% \item Berkeley growth curves in \cite{ChenReiss2014}.
% \item Tecator and phoneme in \cite{Galeano2015} Mahalanobis technometrics paper.
% \item yeast cell cycle gene expression (can't find this publicly) in \cite{LengMuller2005} "Classification using functional data analysis for temporal gene expression data"
% \end{itemize}
% 
% Datasets used in functional manifold papers
% 
% \begin{itemize}
% \item Berkeley growth, yeast cell cycle gene expression (can't find this publicly) in \cite{ChenMuller2012}
% \item Tecator in \cite{LinYao2017} contamination paper
% \item Berkeley growth, gait cycle in \cite{Dimeglio2014} robust isomap paper
% \end{itemize}

### 2019-09-03 index.Rmd
wrote section classification of real data

Let $X_1,\ldots,X_n$ be a sample of $n$ independent realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. Suppose additionally that the function $X$ belongs to a Riemannian manifold $\M \subset L^2([a,b],\R)$ with Riemannian metric tensor $g$ which can be used to assign a metric on the manifold as follows \cite{Lin2014}. For each point $X$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_X$ on the tangent space $T_X \M$. The norm of a tangent vector $V \in T_X \M$ is defined as $$||V|| = \sqrt{g_X(V,V)}.$$ The geodesic distance between two functions $X_i,X_j$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_{\M}(X_i,X_j):=\inf_{\gamma:[0,1] \to \M, \gamma(0) = X_i, \gamma(1) = X_j} l(\gamma)$$
where $$ l(\gamma) := \int_0^1 || \frac{\,d\gamma}{\,d t}(t) || \,dt$$
is the length of the piecewise-smooth curve $\gamma$. 
<!-- ???Is it clear how $l(\gamma)$ depends on $g$??? -->

@@

1. Construct a weighted graph $G$ with nodes corresponding to the observations $x_1,\ldots,x_n\in \R^D$. Two nodes $x_i$ and $x_j$ are connected by an edge $e_{ij}$ with weight $d_{ij} 1\{d_{ij} \le \epsilon \}$ where $d_{ij}=||x_i-x_j ||_{2}$ and $\epsilon > 0$.
2. Estimate the pairwise geodesic distances $d_\M(x_i,x_j)$ based on $G$ using shortest-path algorithms. Specifically, this is the sum of the weights of the edges forming the shortest path, which is calculated either with the Floyd-Warshall algorithm \cite{Floyd1962} or with the Dijkstra algorithm \cite{Dijkstra59anote}.
3. Use multidimensional scaling to obtain an embedding in $\R^d$ that preserves the pairwise geodesic distances estimated above.

@@

In what follows, we call IsoGeo the two-step procedure which performs a modified version of the first step of ISOMAP followed by the original second step of ISOMAP. The first step of IsoGeo constructs the complete weighted ???Marie: does this mean all weights are nonzero??? graph $G_c$ according to \cite{Dimeglio2014} so that the graph is independent of the tuning parameter $\epsilon$. Next, we obtain the minimal spanning tree associated with $G_c$ and denote its set of edges by $E_s$. The graph $G$ is obtained by adding all edges $e_{ij}$ to $E_s$ for which the following condition is true:
$$ \overline{x_ix_j} \subset \bigcup_{i=1}^n B(x_i,\epsilon_i),$$
where $B(x_i,\epsilon_i)$ is the open ball of center $x_i$ and radius $\epsilon_i = \max_{e_{ij}\in E_s}d_{ij}$, and $\overline{x_ix_j}= \{x\in R^D \ | \ \exists \lambda \in [0,1], x=\lambda x_i + (1-\lambda)x_j \}$. Let the distance estimated by IsoGeo be denoted $\text{IsoGeo}(x_i,x_j)$.
Note that IsoGeo has no tuning parameters.

@@

There are two tuning parameters to our procedure. First, there is the dimension $s$ in step 1. In our simulations we try $s \in \{1,2,3\}$. The other parameter is the bandwidth $h$ in the subspace constrained mean-shift. In our simulations we set $h$ according to the heuristic in Equation (A1) of \cite{ChenHo2015}. In reality, depending on the downstream task, both $s$ and $h$ can be tuned in a data-adaptive way, say using cross-validation.

@@

We ran P-ISOMAP separately in Matlab and tuned for the number of neighbors and the penalty parameter using the relative Frobenius assessment metric. We found that the penalty parameter chosen in this oracle fashion was always zero, thus reducing the P-ISOMAP procedure to standard ISOMAP. Given this, we did not make further comparison to P-ISOMAP in our simulation study. We also do not present the results for the RD method which always produced poor estimates of the geodesic distance.

@@

The second metric $||G-\hat G||/||G||$ is the Frobenius norm of the estimation error matrix $G-\hat G$, relative to the Frobenius norm of the true distance matrix $G$. The third metric is the Pearson correlation coefficient between the upper diagonal of $G$ and $\hat G$. A good geodesic distance estimator has low Frobenius norm and high Pearson correlation coefficient.

We shall consider three functional manifolds in our simulation study. Throughout, we make use of a helpful property of one-dimensional manifolds. If $X \in \M$, with the $L^2$ metric tensor, depends on a parameter $\theta \in \mathbb R$, then 
$$
d_\M(X_{\theta_1},X_{\theta_2}) = \int_{\theta_1}^{\theta_2} || \frac{\partial X}{\partial \theta}(t)||_{L^2} \,d\theta.
$$

@@

## Isometric functional manifold of normal density functions
<!-- Scenario 2 in sim_functional_data -->
We take Manifold 2 in \cite{ChenMuller2012} and set the variance of the normal density to be $1$. Let $X_\beta: [a,b] \to \R$ be given by $X_\beta(t) = \frac{1}{\sqrt{2\pi}} \exp{[-\frac{1}{2}(t-\beta)^2]}$. Consider $$\M = \left \{X_\beta: \beta \in [-1,1], t \in [a,b]\right \}$$ with the $L^2$ inner product as the metric tensor. First, note that the "straight" line connecting $X_{\beta_1}$ and $X_{\beta_2}$ does not always stay inside of $\M$. Thus the geodesic distance will not coincide with the $L^2$ distance.

We set $a=-4$ and $b=4$ and sample $\beta$ according to a truncated standard normal with support on $[-1,1]$. The geodesic distance between the curves $X_{\beta_1}$ and $X_{\beta_2}$ is given by
\begin{eqnarray*}
d_\M(X_{\beta_1},X_{\beta_2}) &=& \int_{\beta_1}^{\beta_2} \left \| \frac{d X_\beta (t)}{d\beta} \right\|_{L^2} d\beta \\
% &=&  \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4 \frac{1}{\sqrt{\pi}} \exp\{-(t-\beta)^2\}(t-\beta)^2 dt  }  \  d\beta \\ 
&=&   \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4(t-\beta)^2 f(t) dt  }  \  d\beta   \\
&\approx&  \int_{\beta_1}^{\beta_2}  \sqrt{ \frac{1}{2\sqrt{\pi}} \frac{1}{2}} \ d\beta \\
&=& (\beta_2-\beta_1) \frac{1}{2\pi^{1/4}}
\end{eqnarray*}
\noindent where $f$ is the density of a N$(\beta,1/2)$ and the approximation in the next to last step comes from integrating on $[a,b]=[-4,4]$ rather than $\R$. 
We can see this manifold is isometric, since the geodesic distance between $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ is proportional to $\beta_2-\beta_1$.

## Functional manifold of square root velocity functions
<!-- Scenario 4 in sim_functional_data -->
It was shown in \cite{Joshi2007} that the square root representation of probability density functions gives rise to a simple closed-form geodesic distance.
They consider the manifold $$ \M = \{ \psi:[0,1] \to \R : \psi \ge 0, \int_0^1 \psi^2(s) \,ds = 1 \}$$ with the Fisher-Rao metric tensor $$ <v_1,v_2> = \int_0^1 v_1(s) v_2(s) \,ds $$ for two tangent vectors $v_1,v_2 \in T_\psi(\M)$. Note that this coincides with the $L^2[0,1]$ inner product. It was shown in \cite{Joshi2007} that the geodesic distance between any two $\psi_1$ and $\psi_2$ in $\M$ is simply $$d_\M(\psi_1,\psi_2) = \cos^{-1}<\psi_1,\psi_2>.$$
We will specifically examine the square root of $Beta(\alpha,\beta)$ densities, which are of course supported on $[0,1]$. That is, consider $$ M = \{ \sqrt{\psi_{\alpha,\beta}}: 1 \le \alpha \le 5, 2 \le \beta \le 5\} $$ where $\psi_{\alpha,\beta}: [0,1] \to \R$ is the probability density function of a Beta random variable with shape parameters $\alpha$ and $\beta$. We sample $\alpha$ according to a truncated normal with mean 3 and variance 0.09 with support on $[1,5]$. We sample $\beta$ according to a truncated normal with mean 3.5 and variance 0.09 with support on $[2,5]$.

## Functional manifold of warped functions
<!-- Scenario 5 in sim_functional_data -->
This is based on Equations 17 and 18 in \cite{Kneip2008} but with $z_{i1}, z_{i2}$ both set to $1$. (Note that Equation 17 therein has a typo where the exponentials are missing negative signs.) Let $X_\alpha(t) = \mu(h_\alpha(t))$ be defined on $[-3,3]$ where $$ \mu(t) = \exp\{(t-1.5)^2/2\}+\exp\{(t+1.5)^2/2\}$$ and $$ h_\alpha(t) = 6\frac{\exp\{\alpha(t+3)/6\}-1}{\exp\{\alpha\}-1}$$ if $\alpha \ne 0$ and $h_\alpha(t) = t$ otherwise.

Consider the manifold $$ M = \{X_\alpha: -1 \le \alpha \le 1\}.$$ 
We sample $\alpha$ according to a truncated standard normal with support on $[-1,1]$.
The geodesic distance is then $$ d_\M(X_{\alpha_1},X_{\alpha_2}) = \int_{\alpha_1}^{\alpha_2} \left \| \frac{d X_\alpha (t)}{d\alpha} \right\|_{L^2} d\alpha.$$ We estimate this using numerical integration.

@@

The tuning parameters for the proposed method were chosen as described at the end of the methodology section. In step 1 of our procedure, we take $D = 100$ for the size of the common time grid. We will investigate how two parameters affect the performance of the various methods in consideration. The first one is the signal-to-noise ratio ???[Marie] describe??? which we fix at either $0.1$ (low) or $0.5$ (high). The second parameter is $K$, the dimension of the discretely-observed noisy $Y_i$'s. We set $K$ to either $30$ (sparse) or $100$ (abundant).

We will present the results for the high signal-to-noise and sparse $K$ setting. We show the performance of the Spline IsoGeo and Spline L2 method versus our method for three dimensions $s=1,2,3$. We present the result of the square root velocity manifold scenario in the Appendix. We can see from Figures \ref{fig:normal} and \ref{fig:warping} that spline smoothing followed by either IsoGeo or the $L^2$ distance calculation estimates the pairwise geodesic distances reasonably well though not as well as the proposed method for $s=2$ and $s=3$. Interestingly, when $s$ in our method is not selected properly, for instance when $s=1$, our estimator performs poorly in terms of the relative Frobenius norm although still better than Spline IsoGeo and Spline L2 in the other two metrics. It may be possible that for certain downstream tasks such as clustering or classification, precise estimation of the geodesic distance is less important than one that respects the ordering of the distances. In such cases, the near-isometry and Pearson correlation metrics may give a better indication of downstream performance of the geodesic distance estimator.

@@

>>>>

Let $X_1,\ldots,X_n$ be a sample of $n$ independent realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. Suppose additionally that the function $X$ belongs to a Riemannian manifold $\M \subset L^2([a,b],\R)$ with Riemannian metric tensor $g$ which can be used to assign a metric on the manifold as follows \cite{Lin2014}. For each point $X$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_X$ on the tangent space $T_X \M$. The norm of a tangent vector $V \in T_X \M$ is defined as $$\|V\| = \sqrt{g_X(V,V)}.$$ The geodesic distance between two functions $X_i,X_j$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_{\M}(X_i,X_j):=\inf_{\gamma:[0,1] \to \M, \gamma(0) = X_i, \gamma(1) = X_j} l(\gamma),$$
where $$ l(\gamma) := \int_0^1 \left\| \frac{\,d\gamma}{\,d t}(t) \right\| \,dt$$
is the length of the piecewise-smooth curve $\gamma$. 
<!-- ???Is it clear how $l(\gamma)$ depends on $g$??? -->

@@

1. Construct a weighted graph $\mathcal{G}$ with nodes corresponding to the observations $x_1,\ldots,x_n\in \R^D$. Two nodes $x_i$ and $x_j$ are connected by an edge $e_{ij}$ with weight $d_{ij} 1\{d_{ij} \le \varepsilon \}$ where $d_{ij}=\|x_i-x_j \|_{2}$ and $\varepsilon > 0$.
2. Estimate the pairwise geodesic distances $d_\M(x_i,x_j)$ based on $\mathcal{G}$ using shortest-path algorithms. Specifically, this is the sum of the weights of the edges forming the shortest path, which is calculated either with the Floyd-Warshall algorithm \cite{Floyd1962} or with the Dijkstra algorithm \cite{Dijkstra59anote}.
3. Use multidimensional scaling to obtain an embedding in $\R^d$ that preserves the pairwise geodesic distances estimated above.

@@

In what follows, we call IsoGeo the two-step procedure which performs a modified version of the first step of ISOMAP followed by the original second step of ISOMAP. The first step of IsoGeo constructs a weighted graph $\mathcal{G}$ according to \cite{Dimeglio2014}. First we construct a complete weighted graph $\mathcal{G}_c$, meaning that every pair of nodes $x_i$ and $x_j$ are connected by an edge of weight $d_{ij}$, so that the graph is independent of the tuning parameter $\varepsilon$. Next, we obtain the minimal spanning tree associated with $\mathcal{G}_c$ and denote its set of edges by $E_s$. The graph $\mathcal{G}$ is obtained by adding all edges $e_{ij}$ to $E_s$ for which the following condition is true:
$$ \overline{x_ix_j} \subset \bigcup_{i=1}^n B(x_i,\varepsilon_i),$$
where $B(x_i,\varepsilon_i)$ is the open ball of center $x_i$ and radius $\epsilon_i = \max_{e_{ij}\in E_s}d_{ij}$, and $\overline{x_ix_j}= \{x\in R^D \ | \ \exists \lambda \in [0,1], x=\lambda x_i + (1-\lambda)x_j \}$. Let the distance estimated by IsoGeo be denoted $\text{IsoGeo}(x_i,x_j)$.
Note that IsoGeo has no tuning parameters.

@@

There are two tuning parameters to our procedure. First, there is the dimension $s$ in step 1. In our simulations we try $s \in \{1,2,3\}$. The other parameter is the bandwidth $h$ in the subspace constrained mean-shift. In our simulations we set $h$ according to the heuristic in Equation (A1) of \cite{ChenHo2015}: 
$$h = A_0\left(\frac{1}{s+2}\right)^{\frac{1}{s+4}}n^{\frac{-1}{s+4}}\sigma_{\min}, $$ where $A_0 = ?$ and $\sigma_{\min}$ is the minimum over $j=1,\ldots,s$ of the standard deviation of $\tilde X^s_{1,j},\ldots, \tilde X^s_{n,j}$, with $\tilde X^s_{i,j}$ being the $j$th entry of the vector $\tilde X^s_{i}$. In reality, depending on the downstream task, both $s$ and $h$ can be tuned in a data-adaptive way, say using cross-validation.

@@

We ran P-ISOMAP separately in Matlab and tuned for the parameter $\varepsilon$ and the penalty parameter using the relative Frobenius assessment metric. We found that the penalty parameter chosen in this oracle fashion was always zero, thus reducing the P-ISOMAP procedure to standard ISOMAP. Given this, we did not make further comparison to P-ISOMAP in our simulation study. We also do not present the results for the RD method which always produced poor estimates of the geodesic distance.

@@

The second metric $\|G-\hat G\|/\|G\|$ is the Frobenius norm of the estimation error matrix $G-\hat G$, relative to the Frobenius norm of the true distance matrix $G$. The third metric is the Pearson correlation coefficient between the upper diagonal of $G$ and $\hat G$. A good geodesic distance estimator has low Frobenius norm and high Pearson correlation coefficient.

We shall consider three functional manifolds in our simulation study. Throughout, we make use of a helpful property of one-dimensional manifolds. If $X \in \M$, with the $L^2$ metric tensor, depends on a parameter $\theta \in \mathbb R$, then 
$$
d_\M(X_{\theta_1},X_{\theta_2}) = \int_{\theta_1}^{\theta_2} \left\| \frac{\partial X}{\partial \theta}(t)\right\|_{L^2} \,d\theta.
$$

@@

## Isometric functional manifold of normal density functions
<!-- Scenario 2 in sim_functional_data -->
We take Manifold 2 in \cite{ChenMuller2012} and set the variance of the normal density to be $1$. Let $X_\beta: [a,b] \to \R$ be given by $X_\beta(t) = \frac{1}{\sqrt{2\pi}} \exp{[-\frac{1}{2}(t-\beta)^2]}$. Consider $$\M = \left \{X_\beta: \beta \in [-1,1], t \in [a,b]\right \},$$ with the $L^2$ inner product as the metric tensor. First, note that the "straight" line connecting $X_{\beta_1}$ and $X_{\beta_2}$ does not always stay inside of $\M$. Thus the geodesic distance will not coincide with the $L^2$ distance.

We set $a=-4$ and $b=4$ and sample $\beta$ according to a truncated standard normal with support on $[-1,1]$. The geodesic distance between the curves $X_{\beta_1}$ and $X_{\beta_2}$ is given by
\begin{eqnarray*}
d_\M(X_{\beta_1},X_{\beta_2}) &=& \int_{\beta_1}^{\beta_2} \left \| \frac{d X_\beta (t)}{d\beta} \right\|_{L^2} d\beta \\
% &=&  \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4 \frac{1}{\sqrt{\pi}} \exp\{-(t-\beta)^2\}(t-\beta)^2 dt  }  \  d\beta \\ 
&=&   \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4(t-\beta)^2 f(t) dt  }  \  d\beta   \\
&\approx&  \int_{\beta_1}^{\beta_2}  \sqrt{ \frac{1}{2\sqrt{\pi}} \frac{1}{2}} \ d\beta \\
&=& (\beta_2-\beta_1) \frac{1}{2\pi^{1/4}},
\end{eqnarray*}
\noindent where $f$ is the density of a N$(\beta,1/2)$ and the approximation in the next to last step comes from integrating on $[a,b]=[-4,4]$ rather than $\R$. 
We can see this manifold is isometric, since the geodesic distance between $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ is proportional to $\beta_2-\beta_1$.

## Functional manifold of square root velocity functions
<!-- Scenario 4 in sim_functional_data -->
It was shown in \cite{Joshi2007} that the square root representation of probability density functions gives rise to a simple closed-form geodesic distance.
They consider the manifold $$ \M = \{ \psi:[0,1] \to \R : \psi \ge 0, \int_0^1 \psi^2(s) \,ds = 1 \},$$ with the Fisher-Rao metric tensor $$ <v_1,v_2> = \int_0^1 v_1(s) v_2(s) \,ds ,$$ for two tangent vectors $v_1,v_2 \in T_\psi(\M)$. Note that this coincides with the $L^2[0,1]$ inner product. It was shown in \cite{Joshi2007} that the geodesic distance between any two $\psi_1$ and $\psi_2$ in $\M$ is simply $$d_\M(\psi_1,\psi_2) = \cos^{-1}<\psi_1,\psi_2>.$$
We will specifically examine the square root of $Beta(\alpha,\beta)$ densities, which are of course supported on $[0,1]$. That is, consider $$ \M = \big\{ \sqrt{\psi_{\alpha,\beta}}: 1 \le \alpha \le 5, 2 \le \beta \le 5\big\}, $$ where $\psi_{\alpha,\beta}: [0,1] \to \R$ is the probability density function of a Beta random variable with shape parameters $\alpha$ and $\beta$. We sample $\alpha$ according to a truncated normal with mean 3 and variance 0.09 with support on $[1,5]$. We sample $\beta$ according to a truncated normal with mean 3.5 and variance 0.09 with support on $[2,5]$.

## Functional manifold of warped functions
<!-- Scenario 5 in sim_functional_data -->
This is based on Equations 17 and 18 in \cite{Kneip2008} but with $z_{i1}, z_{i2}$ both set to $1$. (Note that Equation 17 therein has a typo where the exponentials are missing negative signs.) Let $X_\alpha(t) = \mu(h_\alpha(t))$ be defined on $[-3,3]$ where $$ \mu(t) = \exp\{(t-1.5)^2/2\}+\exp\{(t+1.5)^2/2\},$$ and $$ h_\alpha(t) = 6\frac{\exp\{\alpha(t+3)/6\}-1}{\exp\{\alpha\}-1},$$ if $\alpha \ne 0$ and $h_\alpha(t) = t$ otherwise.

Consider the manifold $$ \M = \{X_\alpha: -1 \le \alpha \le 1\}.$$ 
We sample $\alpha$ according to a truncated standard normal with support on $[-1,1]$.
The geodesic distance is then $$ d_\M(X_{\alpha_1},X_{\alpha_2}) = \int_{\alpha_1}^{\alpha_2} \left \| \frac{d X_\alpha (t)}{d\alpha} \right\|_{L^2} d\alpha.$$ We estimate this using numerical integration.

@@

For each manifold $\M$ described above, we simulate ???Susan how many? 1000?? samples of $200$ functions $X_i\in\M$. We then construct vectors $Y_i\in \R^K$ such that $Y_{ij}=X_i(t_j) + \epsilon_{ij}$ with $T=(t_1,\ldots,t_K)$ a common grid of equally spaced points and $\epsilon_{ij}\sim N(0,\sigma_\epsilon^2)$.
The tuning parameters for the proposed method were chosen as described at the end of the methodology section.  In step 1 of our procedure, we take $D = 100$ for the size of the common time grid. We will investigate how two parameters affect the performance of the various methods in consideration. The first one is the signal-to-noise ratio 
$$SNR = 10 log_{10}\left(\frac{\sigma^2_{sig}}{\sigma^2_\epsilon}\right),$$
where $\sigma^2_{sig}$ is the variance of the signal. We fix SNR at either $0.1$ (low) or $0.5$ (high). The second parameter is $K$, the dimension of the discretely-observed noisy $Y_i$'s. We set $K$ to either $30$ (sparse) or $100$ (abundant).

We will present the results for the high signal-to-noise and sparse $K$ setting. We show the performance of the Spline IsoGeo and Spline L2 method versus our method for three dimensions $s=1,2,3$. We can see from Figures \ref{fig:normal} and \ref{fig:warping} that spline smoothing followed by either IsoGeo or the $L^2$ distance calculation estimates the pairwise geodesic distances reasonably well though not as well as the proposed method for $s=2$ and $s=3$. Interestingly, when $s$ in our method is not selected properly, for instance when $s=1$, our estimator performs poorly in terms of the relative Frobenius norm although still better than Spline IsoGeo and Spline L2 in the other two metrics. It may be possible that for certain downstream tasks such as clustering or classification, precise estimation of the geodesic distance is less important than one that respects the ordering of the distances. In such cases, the near-isometry and Pearson correlation metrics may give a better indication of downstream performance of the geodesic distance estimator. The result of the square root velocity manifold scenario is presented in
Figure \ref{fig:betas}. The Spline L2 method has similar performance to
our proposed method. This could be due to the fact that we did not
sample the manifold broadly enough so that the sample lives in a locally
flat region where the \(L^2\) distance holds.

@@

\begin{figure}
\includegraphics[height=0.3\textheight]{sim_results/sce=4_SNR=high_Kobs=30_MSE}
\includegraphics[height=0.3\textheight]{sim_results/sce=4_SNR=high_Kobs=30_isometry}
\includegraphics[height=0.3\textheight]{sim_results/sce=4_SNR=high_Kobs=30_Pearson}
\caption{Manifold of square root Beta densities. Besides Spline IsoGeo, all methods perform comparably well.}
\label{fig:betas}
\end{figure}
